Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-10 _tex/paper2.tex
Built site for gh-pages

\section{Updating CJ's theoretical and statistical
model}\label{sec-theory}

This section presents a theoretical model for CJ that extends
Thurstone's theory. It uses causal inference and, in particular, the
structural approach where structural causal models (SCMs) and directed
acyclic graphs (DAGs) articulate the core theoretical principles of CJ
theory. The model also incorporates several assessment design features
that influence CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition, the study uses Bayesian statistical
methods to translate these theoretical and practical elements into a
statistical model that facilitates the analysis of pairwise comparison
data.

@@

@@

structural causal models (SCMs) and directed acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}

causal analysis \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}

@@

\subsubsection{The assessment design
features}\label{sec-theory-theoretical_AD}

@@

This section introduces fundamental statistical and causal inference
concepts necessary for understanding the core theoretical principles
described in Section~\ref{sec-theory}. It does not, however, offer a
comprehensive overview of statistical and causal inference methods.
Readers seeking more in-depth understanding may wish to explore
introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
\citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}. They may also find
it helpful to consult introductory books like \citet{Pearl_et_al_2018},
\citet{Neal_2020}, and \citet{McElreath_2020}. For more advanced study,
readers may refer to seminal intermediate papers such as
\citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
\citet{Hernan_et_al_2020}.

@@

However, researchers often seek to uncover the mechanisms underlying
specific data and establish causal relationships rather than simply
identify associations. In the example, researchers can interpret the
associational estimate represented by the Z-statistic as causal. This
interpretation relies on the data meeting the assumptions of the Z-test
and the data being collected through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers to interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key properties, such as
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from other variables \(X\) and the outcome \(Y\)
\citep{Morgan_et_al_2014, Neal_2020}.

@@

\emph{Counterfactuals} represent hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. Although a comprehensive
discussion of causes and counterfactuals is beyond the scope of this
document, it is possible to provide a brief overview of how the
framework addresses the fundamental problem of causality. Using the
example outlined in Section~\ref{sec-appendix-B}, the framework begins
by defining the \emph{individual causal effect} (ICE) as the difference
between the students' potential outcomes:
\(\tau_{i} = Y^{1}_{i} - Y^{2}_{i}\). Here, \(Y^{1}_{i}\) represents the
potential outcome observed under \(T_{i}=1\), while \(Y^{2}_{i}\)
represents the potential outcome observed under \(T_{i}=2\). Note that
when a student is assigned to \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) is no longer observed nor observable, making it a
counterfactual. To address the challenge posed by these counterfactuals,
the framework extends the ICE to the \emph{average causal effect} (ACE).
Researchers define the ACE as
\(\tau = E[\tau_{i}] = E[Y^{1}_{i}]- E[Y^{2}_{i}]\), which represents
the difference between the average of observed potential outcomes and
counterfactuals.

Even when data originates from an observational study, researchers can
identify the ACE from associational estimates through causal inference.
They achieve this by performing statistical conditioning on a
\emph{sufficient adjustment set} of variables \(X\)
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially an empty set) should block all (and
not open any) non-causal paths between \(T\) to \(Y\), thereby ensuring
that the estimator for the ACE satisfies several key properties,
including eliminating confounding. If such a set exists, researchers can
express the ACE from associational random quantities as
\(\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
where \(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}. Of course, the validity of the claims about
the effects of \(T\) on \(Y\) now hinges on the assumption that \(X\)
serves as a sufficient adjustment set. However, as
\citet[pp.~150]{Kohler_et_al_2019} noted, ``statements about the real
world from observed (not observational) data require assumptions (and)
this is true no matter what kind of data we have''. For example, if
researchers cannot conduct the randomized experiments described in
Section~\ref{sec-appendix-A} and must rely on observational data, they
can still estimate \(\tau\) if a variable \(X\), such as the
socio-economic status of the school, blocks all (and not open any)
non-causal paths from the teaching method \(T\) to the outcome \(Y\).

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches use rigorous
mathematical notation to characterize the ACE, \emph{not only for
dichotomous but also for continuous variables} \(T\), though they do so
in different ways \citep{Neal_2020}. The potential outcomes approach
relies on counterfactual notation, as seen in the above examples,
whereas the structural approach uses do-calculus \citep{Pearl_2009}.
Despite these differences, both notations can be expressed in terms of
the other, and both approaches provide methods for using experimental
and observational data to identify and estimate causal effects
\citep{Pearl_2010}.

@@

The structural approach to causal inference, however, offers a notable
advantage over the potential outcomes approach. It enables researchers
to formally and graphically represent a system using SCMs and DAGs
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. SCMs
and DAGs function as conceptual models describing the presumed causal
structure underlying the ACE. In essence, these models act as
\emph{theoretical models} that guide researchers in determining which
statistical models can yield valid causal inferences, assuming the
depicted causal structure of the models is correct
\citep{McElreath_2020}. Notably, every SCM has a associated DAG
\citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a visual
representation of the role of theoretical models in the inference
process.

SCMs and DAGs offer two key advantages for modeling causal structures.
First, they enable researchers to represent causal relations in a
non-parametric and fully interactive manner. This feature allows for
feasible ACE identification strategies without specifying the data type
or the nature of the functional dependence among variables
\citep{Morgan_et_al_2014}. Second, regardless of complexity, they can
represent a wide range of causal structures using only five fundamental
building blocks \citep{Neal_2020, McElreath_2020}. This feature allows
researchers to decompose complex structures, facilitating their analysis
by focusing on the causal assumptions associated with each building
block \citep{McElreath_2020}.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display the five fundamental building blocks of SCMs and DAGs. The left
panels of the figures show the formal mathematical models, represented
by the SCMs, defined in terms of a set of \emph{endogenous} variables
\(V=\{X,Z,Y\}\), a set of \emph{exogenous} variables
\(E=\{e_{X},e_{Z},e_{Y}\}\), and a set of functions
\(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
are those whose causal mechanisms a researchers chooses to model
\citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

The right panels of the figures display the complementary DAGs. A DAG
consists of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} means the edges extend from one node
to another, with arrows indicating the direction of causal influence.
The term \emph{acyclic} signifies that the causal influences do not form
loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
expected direction of causal influences among these variables. Often,
DAGs omit the exogenous variables for simplicity (the \emph{standard}
representation). However, including exogenous variables in a graph (the
\emph{magnified} representation) can be beneficial, as their presence
can reveal potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

@@

Researchers can use these building blocks to depict the scenario
described in Section~\ref{sec-appendix-B}. In this scenario, they rely
on observational data because it is not feasible to run a randomized
experiment. Furthermore, the scenario indicates that researchers have
measured a variable \(X\) (the socio-economic status of the school) that
is assumed to block all non-causal paths from the teaching method \(T\)
to the outcome \(Y\). Figures \ref{fig-scm_example1} and
\ref{fig-mdag_example1} illustrate the plausible causal structure for
this example. The figures highlight the presence of at least four of the
five fundamental building blocks. The figures display multiple
descendants, as shown by pairwise relationships such as
\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
also feature multiple pairs of unconnected nodes, evident from the
relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
Finally, the figures illustrate one fork, \(X \rightarrow \{T, Y\}\),
and two colliders: \(\{X, e_{T}\} \rightarrow T\) and
\(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example}Plausible causal structure for the example.}

>>>>

\section{Updating the theoretical and statistical
model}\label{sec-theory}

This section uses the structural approach to causal inference
\citep{Pearl_2009, Pearl_et_al_2016} to articulate a theoretical model
that captures the core principles of Thurstone's theory. The model also
incorporates various assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Finally, the section employs Bayesian statistical methods to transform
these theoretical and practical elements into a statistical model that
facilitates the analysis of pairwise comparison data. See
Section~\ref{sec-appendix} for an overview of the statistical and causal
inference concepts required for the development of this section.

@@

The theoretical model uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of the
CJ system. First, \emph{the population model} is created to represent a
conceptual population of CJ experiments. The model then integrates
various assessment design features relevant to CJ experiments, leading
to the development of \emph{the sample-comparison model}.

@@

@@

\subsubsection{The sample-comparison
model}\label{sec-theory-theoretical_SC}

@@

This section introduces fundamental statistical and causal inference
concepts necessary for understanding the core theoretical principles
described in Section~\ref{sec-theory}. It does not, however, offer a
comprehensive overview of causal inference methods. Readers seeking more
in-depth understanding may wish to explore introductory papers such as
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find it helpful to consult
introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
\citet{McElreath_2020}. For more advanced study, readers may refer to
seminal intermediate papers such as \citet{Neyman_et_al_1923},
\citet{Rubin_1974}, \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009},
as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
and \citet{Hernan_et_al_2020}.

@@

However, researchers often seek to uncover the mechanisms underlying
specific data and establish causal relationships rather than simply
estimate associations. In the example, researchers can interpret the
associational estimate represented by the Z-statistic as causal. This
interpretation relies on the data meeting the assumptions of the Z-test
and being collected through a randomized experiment.

Randomized experiments are widely recognized as the gold standard in
evidence-based science \citep{Hariton_et_al_2018, Hansson_2014}. This
recognition stems from their ability to enable researchers to interpret
associational estimates as causal. They achieve this by ensuring data,
and by extension an estimator, satisfies several key properties, such as
common support, no interference, and consistency
\citep{Morgan_et_al_2014, Neal_2020}. The most critical property,
however, is the elimination of confounding. \emph{Confounding} occurs
when an external variable \(X\) simultaneously influences the outcome
\(Y\) and the variable of interest \(T\), resulting in spurious
associations \citep{Everitt_et_al_2010}. Randomization addresses this
issue by decoupling the association between the intervention allocation
\(T\) from any other variable \(X\)
\citep{Morgan_et_al_2014, Neal_2020}.

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendix-A}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes:
\(\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)\). Here,
\(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i}|do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i}|do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address the challenge of unobserved counterfactuals, the
approach extends the ICE to the \emph{average causal effect} (ACE):
\(\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]\),
representing the average difference between observed potential outcomes
and their counterfactual counterparts. Notably, the approach extends the
ACEs for categorical and continuous variables \(T\), ensuring broad
applicability across different causal scenarios.

Even when data originates from an observational study, researchers can
identify the ACE from associational estimates using the structural
approach. They achieve this by performing statistical conditioning on a
\emph{sufficient adjustment set} of variables \(X\)
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. This
\emph{sufficient} set (potentially empty) must block all non-causal
paths between \(T\) to \(Y\) without opening new ones, ensuring the ACE
estimator satisfies several key properties, including confounding
elimination. If such a set exists, then \(T\) and \(Y\) are
\emph{d-separated} by \(X\) \citep{Pearl_2009}, meaning researchers can
estimate the ACE from associational random quantities
\citep{Morgan_et_al_2014}. Naturally, the validity of claims about the
effects of \(T\) on \(Y\) now hinges on the assumption that \(X\) serves
as a sufficient adjustment set. However, as
\citet[pp.~150]{Kohler_et_al_2019} observed, drawing conclusions about
the real world from observed data inevitably requires assumptions. This
holds true for both observational and experimental data.

For instance, if researchers are unable to conduct the randomized
experiments described in Section~\ref{sec-appendix-A} and instead rely
on observational data, they can still estimate the ACE provided a
variable \(X\), such as the socio-economic status of the school, blocks
all non-causal paths between the teaching method \(T\) to the outcome
\(Y\) without opening any new ones. Under these conditions, the ACE can
be estimated from associational quantities as
\(\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}.

@@

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}. In
essence, SCMs and DAGs act as \emph{conceptual models} that guide
researchers in determining which statistical models can yield valid
causal inferences, assuming the depicted causal structure of the models
is correct \citep{McElreath_2020}. Notably, every SCM has an associated
DAG \citep{Cinelli_et_al_2020}. Figure~\ref{fig-IEflow} provides a
visual representation of the role of theoretical models in the inference
process.

SCMs and DAGs offer two key advantages for modeling causal structures.
First, they enable the representation of causal relations in a
non-parametric and fully interactive manner. This feature allows for
feasible ACE identification strategies without specifying the data type
or the nature of the functional dependence among variables
\citep{Morgan_et_al_2014}. Second, regardless of complexity, they can
represent a wide range of causal structures using only five fundamental
building blocks \citep{Neal_2020, McElreath_2020}. This feature allows
researchers to decompose complex structures, facilitating their analysis
by focusing on the causal assumptions associated with each building
block \citep{McElreath_2020}.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(V=\{X,Z,Y\}\), a set of \emph{exogenous} variables
\(E=\{e_{X},e_{Z},e_{Y}\}\), and a set of functions
\(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Cinelli_et_al_2020, Neal_2020}. Endogenous variables
are those whose causal mechanisms a researcher chooses to model
\citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

The right panels of the figures display the complementary DAGs. A DAG
consists of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} means the edges extend from one node
to another, with arrows indicating the direction of causal influence.
The term \emph{acyclic} signifies that the causal influences do not form
loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs show the
expected direction of causal influences among these variables. While
DAGs often omit exogenous variables for simplicity (the \emph{standard}
representation), including these variables is advantageous (the
\emph{magnified} representation shown in the figures), as their
inclusion helps to identify potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

@@

Researchers can use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. Figures \ref{fig-scm_example1}
and \ref{fig-mdag_example1} depict the plausible causal structure for
this example, assuming that the variable \(X\) (socio-economic status of
the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures show several
descendant, such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\), and also highlight multiple pairs of unconnected
nodes, evident from the relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
Additional, the figures depict one fork, \(X \rightarrow \{T, Y\}\), and
two colliders: \(\{X, e_{T}\} \rightarrow T\) and
\(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\caption{\label{fig-example}Plausible causal structure the scenario
outlined in Section~\ref{sec-appendix-B}.}
